Staged Abstract Interpreters

### 2019-09-11 paper/10_intro.tex
polish

Abstract interpretation as a lattice-based approach to sound static analyses was
proposed by \citet{DBLP:conf/popl/CousotC77, CousotCousot79-1}. Based on the
notion of Galois connections, an analyzer can soundly approximate concrete
program behavior at runtime by computing fixed points on an abstract domain.
Despite the tremendous theoretical development of abstract interpretation over
the years, constructing artifacts and analyzers that perform sound abstract
interpretation for modern and expressive languages was considered complicated
for a long time.

@@

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs interpretive
overhead, such as pattern matching on the AST nodes and recursive calls on
subexpressions. If this requires only a single traversal, such overhead can be
negligible. Usually, however, abstract interpreters iteratively analyze and
traverse an AST multiple times to reach fixed-points; the accumulated overhead
is therefore magnified.
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).
Second, it is common that abstract interpreters written in pure languages make
extensive use of functional encodings of side-effects to implement the semantics
of abstract interpretation.  For example, the list monad can be used to model
nondeterminism from collecting interpreters. Although such pure approaches have
their own merits, they are significantly slower when compared with imperative,
stateful implementations.  In order to reconcile these approaches, one must use
a programming technique which allows for high-level abstractions during
implementation, but bypasses the incurred overhead at runtime.

Roughly at the same time abstract interpretation was proposed,
\citet{futamura1971partial} observed a close connection between interpreters and
compilers through a hierarchy of specializations, known since then as the
Futamura projections. The first Futamura projection states that specializing an
interpreter to an input program removes interpretative overhead and yields an
equivalent executable that runs faster. In recent years, the idea of Futamura
projections has attracted more attention, and has been successfully applied to
many real-world problems, including PyPy \cite{Bolz:2009:TMP:1565824.1565827},
Truffle \cite{Marr:2015:TVP:2814270.2814275}, query compilers
\cite{DBLP:conf/sigmod/TahboubER18}, etc. The challenge here: can we apply
Futamura projections to a general abstract interpreter to make it run faster?
The solution is not obvious, as we need to specialize a nonstandard semantics
that extensively uses nondeterminism, operates on abstract domains, and
computes fixed-points iteratively over caches.

In this paper, we show that the first Futamura projection can be naturally
extended to abstract interpreters. We present an abstraction-without-regret
framework that eliminates performance penalties for monadic abstract
interpreters, while keeping the implementation as close to their conceptual
model as possible. In short, we borrow ideas and techniques from multi-stage
programming and embedded DSLs, and apply them to abstract interpreters:  1) To
remove the overhead from interpretation and effect layers, we specialize the
abstract interpreter via staging, and then generate efficient low-level code
that performs the actual analysis. 2) Inspired by the tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic interpreter that
abstracts over binding-times and different semantics, which allows the staged
abstract interpreter to be derived from its unstaged counterpart.  As a result,
the derived staged abstract interpreter has no intrusive changes to the
underlying abstract semantics, thereby preserving soundness.  In this sense, our
approach allows for a complete absence of regret in terms of both performance
and engineering effort. We elaborate these two main ideas in detail, as follows.

@@

The first Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent executable. For
instance, assume that @eval : Expr --> Input --> Value@ is an interpreter for some language.
Given a program $e_0$@ : Expr@, by applying the specialization we can
obtain a specialized program $\texttt{eval}_{\texttt{e0}}$ : @Input --> Value@.
%which is the so-called \textit{equivalent executable}.
% By definition of the interpreter, they produce the same result when applied with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.
The first proposed approach to realizing Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes between compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users to explicitly program
over a two-level syntax \cite{Nielson:1992:TFL:130665} by manually annotating
the binding-times of variables. The MSP system will then check whether these
annotations are consistent, and specialize the program using that information.
The staging annotations can be either syntactic (e.g., quote and quasi-quote in
MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} in Scala).

To apply the Futamura projections to abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract interpreter
as the unstaged baseline.  Similar to two-level semantics \cite{NIELSON1989117},
monads provide a good abstraction to hide the detail of abstract interpretation
semantics. However, monads do not have a clear stage distinction. We further
introduce binding-times and make the monads binding-time polymorphic. Our
proposed framework adopts the type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projection of
an abstract interpreter for a stateful higher-order language. By deriving staged
monads that can be used to perform code generation, we obtain a staged abstract
interpreter that can be used for specialization. Through staging, the generated
code is specialized to the input program, and all the monadic operations are
inlined and compiled down to low-level Scala code.

@@

Although the four artifacts may look dissimilar at first glance, they are in
fact all firmly rooted in the concrete semantics.  This observation provides a
way to abstract over the interpreter and achieve the flexibility of
reinterpreting the shared interpreter.  The generic interpreter returns a value
of monadic types, which vary based on semantics. The value domains
of the interpreter and the effects such as state and nondeterminism can all be
wrapped into this monadic type.  The binding-time abstraction is represented by
a higher-kinded type and injected into the monadic type.  The binding-times
control whether the interpreter produces values directly or generates code.

@@

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical performance evaluation.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change (modulo the addition of stage annotations),
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup.

>>>>

Abstract interpretation as a lattice-based approach to sound static analyses was
proposed by \citet{DBLP:conf/popl/CousotC77, CousotCousot79-1}. Based on the
notion of Galois connections, an analyzer can soundly approximate concrete
program behavior at runtime by computing fixed points on an abstract domain.
Despite the tremendous theoretical development of abstract interpretation over
the years, constructing artifacts and analyzers that perform sound abstract
interpretation for modern and expressive languages is considered complicated
for a long time.

@@

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs interpretive
overhead, such as pattern matching on the AST nodes and recursive calls on
subexpressions. If this requires only a single traversal, such overhead can be
negligible. Usually, however, abstract interpreters iteratively analyze and
traverse an AST multiple times to reach fixed-points; the accumulated overhead
is therefore magnified.
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).
Second, it is common that abstract interpreters written in pure languages make
extensive use of functional encodings of side-effects to implement the semantics
of interpreted languages. For example, the list monad can be used to model
nondeterminism of collecting interpreters. Although such pure approaches have
their merits, they are significantly slower when compared with imperative,
stateful implementations.  In order to reconcile these approaches, one need to
use a programming technique which allows for high-level abstractions during
implementation, but bypasses the incurred overhead at runtime.

Roughly at the same time abstract interpretation was proposed,
\citet{futamura1971partial} observed a close connection between interpreters and
compilers through a hierarchy of specializations, known since then as the
Futamura projections. The first Futamura projection states that specializing an
interpreter to an input program removes interpretative overhead and yields an
equivalent executable that runs faster. In recent years, the idea of Futamura
projections has attracted more attention, and has been successfully applied to
many real-world problems, including PyPy \cite{Bolz:2009:TMP:1565824.1565827},
Truffle \cite{Marr:2015:TVP:2814270.2814275}, query compilers
\cite{DBLP:conf/sigmod/TahboubER18}, etc. The challenge here: can we apply
Futamura projections to a general abstract interpreter to make it run faster?
The solution is not obvious, as we need to specialize a nonstandard semantics
that extensively uses nondeterminism, operates on abstract domains, and
computes fixed-points iteratively.

In this paper, we show that the first Futamura projection can be naturally
extended to abstract interpreters. We present an abstraction-without-regret
framework that eliminates performance penalties for monadic abstract
interpreters, while keeping the implementation as close to their conceptual
model as possible. In short, we borrow ideas and techniques from multi-stage
programming and embedded DSLs, and apply them to abstract interpreters:  1) To
remove the overhead from interpretation and effect layers, we specialize the
abstract interpreter via staging, and then generate efficient low-level code
that performs the actual analysis. 2) Inspired by tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic interpreter that
abstracts over binding-times \cite{Ofenbeck:2017:SGP:3136040.3136060,
Amin:2017:CTI:3177123.3158140} and different semantics, which allows the staged
abstract interpreter to be derived from its unstaged counterpart.  As a result,
the derived staged abstract interpreter has no intrusive changes to the
underlying abstract semantics, thereby preserving soundness.  In this sense, our
approach allows for a complete absence of regret in terms of both performance
and engineering effort. We elaborate these two main ideas in detail, as follows.

@@

The first Futamura projection shows that specializing an
interpreter w.r.t. a input program yields an equivalent executable. For
instance, assume that @eval : Expr --> Input --> Value@ is an interpreter for some language.
Given a program $e_0$@ : Expr@, by applying the specialization we can
obtain a specialized program $\texttt{eval}_{\texttt{e0}}$ : @Input --> Value@.
%which is the so-called \textit{equivalent executable}.
% By definition of the interpreter, they produce the same result when applied with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.
The first proposed approach to realizing Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes between compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users to explicitly program
over a two-level syntax \cite{Nielson:1992:TFL:130665} by manually annotating
the binding-times of variables. The MSP system will then check whether these
annotations are consistent, and specialize the program using that information.
The staging annotations can be either syntactic (e.g., quote and quasi-quote in
MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} in Scala).

To apply the Futamura projections to abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract interpreter
as the unstaged baseline.  Similar to two-level semantics \cite{NIELSON1989117},
monads provide a good abstraction to hide the detail of abstract interpretation
semantics. However, monads do not have a clear stage distinction. We further
introduce binding-times and make the monads binding-time polymorphic. Our
proposed framework adopts type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projections of
an abstract interpreter for a stateful higher-order language. By deriving staged
monads that can be used to perform code generation, we obtain a staged abstract
interpreter that can be used for specialization. Through staging, the generated
code is specialized to the input program, and all the monadic operations are
inlined and compiled down to low-level Scala code.

@@

Although the four artifacts may look dissimilar at first glance, they are in
fact all firmly rooted in the concrete semantics.  This observation provides a
way to abstract over the semantics and achieve the flexibility of
reinterpreting the shared interpreter.  The generic interpreter returns a value
of monadic types, which vary based on semantics. The value domains
of the interpreter and the effects such as state and nondeterminism can all be
wrapped into this monadic type.  The binding-time abstraction is represented by
a higher-kinded type and injected into the monadic type.  The binding-times
control whether the interpreter produces values directly or generates code.

@@

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical performance evaluation.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change (modulo the addition of stage annotations),
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to cover different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup.

### 2019-09-11 paper/20_prelim.tex
polish

In this section, we first describe the abstract syntax of the language for our
interpreters. Then, we present the generic interpreter shared among the four
different semantics, after which we instantiate the interpreter to the concrete
one. It is worth noting that we choose to use Scala and a monadic style to
demonstrate this idea, though the approach is not restricted to this choice.
Indeed, one can use imperative or direct style in other MSP languages (e.g.,
MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} or
Template Haskell \cite{Sheard:2002:TMH:636517.636528}) to construct such staged
abstract interpreters.

@@

We consider a call-by-value $\lambda$-calculus in direct-style, extended with
numbers, arithmetic, recursion, and conditionals. Other effects such as
assignments can also be supported readily. Since we are mostly interested in
analyzing dynamic behavior of programs, we elide the static semantics. We also
assume that input programs are well-typed and all variables are distinct. The
abstract syntax is as follows:
\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursion
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetic
\end{lstlisting}

The abstract syntax we present can be seen as a deep embedding of the language:
we use data-types to represent programs. This design choice allows us to use
different interpretations over the AST; with the inheritance and overriding
mechanisms in Scala, we may also add new language constructs and reuse existing
interpretations.

@@

The Scala compiler translates the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@. The last binding
in @for@-comprehensions is translated into a @map@, where the argument of
@yield@ becomes the body expression inside that @map@ operation. The
bindings before the last one are all translated into calls of @flatMap@:
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

@@

In this section, we present the generic interface in the style of a big-step
definitional interpreter. The key idea is to keep both the binding-time type and
returned monadic type abstract, so that they can be instantiated
differently. We also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \citet{DBLP:conf/popl/LiangHJ95} or abstract interpretation
\cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.

@@

\paragraph{Binding-time Abstraction}
The binding-time type is declared as a higher-kinded type @R[_]@.
Now we can use @R@ to annotate other data-types in the interpreter;
it is also injected into the monadic type @MonadOps@.  If we instantiate
@R@ as an identity type (i.e., @type R[T] = T@), then the generic interpreter is a
standard definitional interpreter that will execute the program.  In Section
\ref{stagedinterp}, we instantiate @R@ using LMS's built-in next-stage
type annotation @Rep@, which makes the interpreter act as a compiler.

@@

Other primitive operations in the interpreter handle the language constructs.  The
methods @num@ and @close@ deal with primitive values, which lift literal terms
(e.g., lambdas) to our value representation (e.g., closures).
Conditionals and arithmetic are handled by @br0@ and @arith@, respectively. The
method @ap_clo@ is used for applying functions, which takes a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are all
annotated by @R[_]@, as they are potentially next-stage values when the
interpreter acts as a compiler.
\begin{lstlisting}
  def num(i: Int): Ans
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(fun: R[Value], arg: R[Value]): Ans
\end{lstlisting}

@@

\paragraph{The Interpreter} We can now construct the semantics-agnostic interpreter
in monadic form, as shown in Figure \ref{fig:shared_int}. The basic idea
of generic interpretation is to traverse the AST
while maintaining the effects (such as environment and state updates).
Note that the interpreter is written in open-recursive style: it cannot
refer to itself directly. Instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method @close@ that
lifts $\lambda$-terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because evaluations may happen inside them.
To close the open recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, there is a top-level wrapper @run@. The return type @Result@
depends on the kind of monads being used and is therefore also left abstract.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

@@

As the first step in our roadmap, we instantiate the generic interpreter for
concrete execution in this section. The result is a standard definitional
interpreter with environments and stores. It can also be obtained by
refunctionalizing a standard CESK machine \cite{Felleisen:1987:CAH:41625.41654,
DBLP:conf/ppdp/AgerBDM03}. We first present the concrete components, i.e., the
value domains, then show the monad stack for concrete interpretation, finally
sketch how the primitive operations are implemented.

@@

\paragraph{Unstaged Monads}
For concrete interpretation, the monad needs to model reader and state effects,
which correspond to the environment and the store, respectively. We follow the
monad transformer approach \cite{DBLP:conf/popl/LiangHJ95},
and use the @ReaderT@ and @StateT@ monad transformers to
compose the monad stack. In other words, the type @AnsM@ is
instantiated by layering the @ReaderT@ and @StateT@ transformers\footnote{The
question mark syntax is a kind projector \cite{kindprojector}, such that
\texttt{StateT[IdM,Store,?]} is equivalent to \newline \texttt{(\{type
M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is parameterized by the
type @Env@, and the @StateT@ is parameterized by the type @Store@, and the
inner-most monad @IdM@ is merely an identity monad.
\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T] // the monad stack
    ... }
\end{lstlisting}

Here, we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the environment type, and @M[_]@ is the inner monadic type.
Given a value of @R@, a @ReaderT@ monad produces a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a value of @S@, a @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (of type @S@) is accompanied with the result (of type @A@).
Note that for the moment, the binding-time type @R@ is an identity type; thus,
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.
\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

@@

\paragraph{Primitive Operations}
Other primitive operations over the value domains can be implemented
straightforwardly. We elide most of them but describe a bit on how we handle
functions and applications. The reason is that $\lambda$-terms are data but are also
part of the control flow. The way we treat them is simple now, but
will become more involved when we stage the abstract interpreter.
The method @close@ denotes a $\lambda$-term to our representation of functions.
At the moment, @close@ takes a $\lambda$-term and an environment, and
produces the defunctionalized representation of closures.
The method @ap_clo@ performs function applications, which takes a function
value @fun@ and an argument @arg@. It first extracts the syntactic
$\lambda$-term and environment enclosed in @fun@, and allocates an addresses
for the argument and updates the environment and store accordingly.  Finally,
body expression of the $\lambda$-term is evaluated under the new environment
and store.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
  def ap_clo(ev: Expr => Ans)(fun: Value, arg: Value): Ans = fun match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → arg)
      rt <- local_env(ev(e))(ρ + (x → α))
    } yield rt
  }
\end{lstlisting}

>>>>

In this section, we first describe the abstract syntax of the language. Then,
we present the generic interpreter shared among the four different semantics,
after which we instantiate the interpreter to the concrete one. It is worth
noting that we choose to use Scala and a monadic style to demonstrate this
idea, though the approach is not restricted to this choice.  Indeed, one can
use imperative or direct-style in other MSP languages (e.g., MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} or Template
Haskell \cite{Sheard:2002:TMH:636517.636528}) to construct such staged abstract
interpreters, although the details are different.

@@

We consider a call-by-value $\lambda$-calculus in direct-style, extended with
numbers, arithmetic, recursion, and conditionals. Other effects such as
assignments can also be supported readily. Since we are interested in
analyzing the dynamic behavior of programs, we elide the static semantics. We also
assume that input programs are well-typed and all variables are distinct. The
abstract syntax is as follows:
\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursion
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetic
\end{lstlisting}

The abstract syntax we present can be seen as a deep embedding of the language:
we use data-types to represent programs. This is the most natural choice
for program analysis and allows us to use different interpretations over the
AST; with the inheritance and overriding mechanisms in Scala, we may also add
new language constructs and reuse existing interpretations.

@@

The Scala compiler translates the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@. The last binding
in @for@-comprehensions is translated into a @map@ operation, where the argument of
@yield@ becomes the body expression inside that @map@ operation. The
bindings before the last one are all translated into calls of @flatMap@:
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

@@

In this section, we present the generic evaluator interface in the style of a big-step
definitional interpreter. The key idea is to keep both the binding-time type and
returned monadic type abstract, so that they can be instantiated
differently. We also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \cite{DBLP:conf/popl/LiangHJ95} or abstract interpretation
\cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.

@@

\paragraph{Binding-time Abstraction}
The binding-time type is declared as a higher-kinded type @R[_]@ \cite{Ofenbeck:2017:SGP:3136040.3136060}.
Now we can use @R@ to annotate other data-types in the interpreter;
it is also injected into the monadic type @MonadOps@.  If we instantiate
@R@ as the identity type (i.e., @type R[T] = T@), then the generic interpreter is a
standard definitional interpreter that will execute the program.  In Section
\ref{stagedinterp}, we instantiate @R@ using LMS's built-in next-stage
type annotation @Rep@, which makes the interpreter act as a compiler.

@@

Other primitive operations in the interpreter handle the language constructs.  The
methods @num@ and @close@ deal with primitive values, which lift literal terms
(e.g., lambdas) to our value representation (e.g., closures).
Conditionals and arithmetic are handled by @br0@ and @arith@, respectively. The
method @ap_clo@ is used for applying functions, by taking a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are all
annotated by @R[_]@, as they are potentially next-stage values when the
interpreter acts as a compiler.
\begin{lstlisting}
  def num(i: Int): Ans
  def close(ev: Expr => Ans)(λ: Lam, ρ: R[Env]): R[Value]
  def br0(test: R[Value], thn: => Ans, els: => Ans): Ans
  def arith(op: Symbol, v1: R[Value], v2: R[Value]): R[Value]
  def ap_clo(ev: Expr => Ans)(fun: R[Value], arg: R[Value]): Ans
\end{lstlisting}

@@

\paragraph{The Interpreter} We can now construct the semantics-agnostic interpreter
in monadic form, as shown in Figure \ref{fig:shared_int}. The basic idea
of generic interpretation is to traverse the AST
while maintaining the effects (such as environment and state updates).
Note that the interpreter is written in open-recursive style: it cannot
refer to itself directly. Instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, method @close@ that
lifts $\lambda$-terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because further evaluation may happen inside them.
To close the open recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, there is a top-level wrapper @run@. The return type @Result@
depends on the kind of monads being used and is therefore also left abstract.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

@@

As the first step in our roadmap, we instantiate the generic interpreter for
concrete execution in this section. The result is a standard definitional
interpreter with environments and stores. It can also be obtained by
refunctionalizing a standard CESK machine \cite{Felleisen:1987:CAH:41625.41654,
DBLP:conf/ppdp/AgerBDM03}. We first present the concrete components, i.e., the
value domains, then show the monad stack for concrete interpretation, and finally
sketch how the primitive operations are implemented.

@@

\paragraph{Unstaged Monads}
For concrete interpretation, the monad needs to model reader and state effects,
which correspond to the environment and the store, respectively. We follow the
monad transformer approach \cite{DBLP:conf/popl/LiangHJ95},
and use the @ReaderT@ and @StateT@ monad transformers to
compose the monad stack. In other words, the type @AnsM@ is
instantiated by layering the @ReaderT@ and @StateT@ transformers\footnote{The
question mark syntax is a kind projector \cite{kindprojector}, such that
\texttt{StateT[IdM,Store,?]} is equivalent to \newline \texttt{(\{type
M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is parameterized by the
type @Env@, and the @StateT@ is parameterized by the type @Store@, and the
inner-most monad @IdM@ is merely the identity monad.
\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T] // the monad stack
    ... }
\end{lstlisting}

Here, we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the environment type, and @M[_]@ is the inner monadic type.
Given a value of @R@, a @ReaderT@ monad produces a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a value of type @S@, a @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (of type @S@) is accompanied with the result (of type @A@).
Note that for the moment, the binding-time type @R@ is the identity type; thus,
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.
\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

@@

\paragraph{Primitive Operations}
Other primitive operations over the value domains can be implemented
straightforwardly. We elide most of them but describe how we handle
functions and applications. The reason is that $\lambda$-terms are data but are also
part of the control flow. The way we treat them is simple now, but
will become more involved when we stage the abstract interpreter.
The method @close@ denotes a $\lambda$-term to our representation of functions.
At the moment, @close@ takes a $\lambda$-term and an environment, and
produces the defunctionalized representation of closures.
The method @ap_clo@ performs function applications, which takes a function
value @fun@ and an argument @arg@. It first extracts the syntactic
$\lambda$-term and environment enclosed in @fun@, and allocates an addresses
for the argument and updates the environment and store accordingly.  Finally,
by recursively calling @ev@, it evaluates the body expression of the
$\lambda$-term under the new environment and store.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
  def ap_clo(ev: Expr => Ans)(fun: Value, arg: Value): Ans = fun match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → arg)
      rt <- local_env(ev(e))(ρ + (x → α))
    } yield rt
  }
\end{lstlisting}
